---
title: "Explanatory data analysis with PCA. Decathlon data analysis"
author: "Julie Josse & Sylvain Le Corff"
date: "10/05/2018"
output:
  pdf_document: default
  html_document: default
---

#  Principal Component Analysis on decathlon data

```{r import-load, eval = TRUE, results='hide'}
# Install the package
# install.packages("FactoMineR", dependencies = TRUE)
# Load the package
library(FactoMineR)
data(decathlon)
head(decathlon)
```

This dataset contains the results of decathlon events during two athletic meetings which took place one month apart in 2004: the Olympic Games in Athens (23 and 24 August), and the Decastar 2004 (25 and 26 September). For both competitions, the following information is available for each athlete: performance for each of the 10 events, total number of points (for each
event, an athlete earns points based on performance; here the sum of points
scored) and final ranking. The events took place in the following order: 100 meters, long jump, shot put, high jump, 400 meters (first day) and 110 meter hurdles, discus, pole vault, javelin, 1500 meters (second
day). Nine athletes participated to both competitions. We would like to obtain a typology of the performance profiles.

1) You should inspect the data set with the following commands:
```{r dim, eval=TRUE}
summary(decathlon)
dim(decathlon)
#View(decathlon)
#?decathlon
rownames(decathlon)
```


2) Explain the interest of centering and scaling the data. Could you spotlight outstanding athletes ? Which inequality are you using?

When the data are standardized, it is possible to compare two variables with different units and to say sentences such as "Paul is more remarkable by his height than John is by is weight".
When looking at the standardized data, we can look for the values greater or smaller than 2 for instance. We are refering to Bienayme-Tchebychev which states that 25% of the observations will be at 2 standard deviation from their means. If we consider Gaussian data, we can refine this inequality and consider now that 4.5 % of the observations are greater than 2 in absolute value. Sebrle value for Javeline is 2.528251350 meaning that he is far above average. 


The aim of conducting PCA on this dataset is to determine profiles for similar performances: are there any athletes who are
better at endurance events or those requiring short bursts of energy, etc? And are some of the events similar? If an athlete performs well in one event, will he necessarily perform well in another?

3) Explain your choices for the active and illustrative variables/individuals and perform the PCA on this data set.

```{r, eval = TRUE, include = TRUE}
 res.pca <- PCA(decathlon, quanti.sup = c(11,12) , quali.sup = 13)
 summary(res.pca, nbelements = 5, ncp = 4)
 barplot(res.pca$eig[,1], main = "Eigenvalues",
 names.arg = paste("Dim",1:nrow(res.pca$eig), sep=""))
 plot(res.pca, choix = "ind", axes = c(3, 4))
 plot(res.pca, choix = "var", axes = c(3, 4))
```


To obtain a typology of the athletes based on their performances for the 10
decathlon events, such as "two athletes are close as they have similar
performance profiles", the distances between two athletes are defined on the
basis of their performances in the 10 events. Thus, only the performance
variables are considered active; the other variables (number of points, rank,
and competition) are supplementary. Here, the athletes are all considered as active
individuals.

4) Comment the percentage of variability explained by the two first dimensions. What would you like (a small percentage, a high percentage) and why?

The first two dimensions  summarize 50\% of the total inertia, i.e. 50\% of the total variability of the cloud of individuals (or variables) is represented by the first 2d plane. The importance of this percentage cannot be assessed without taking into account the number of active individuals and active variables. It may be interesting to compare this percentage with that of the 0.95 -quantile distribution of the percentages obtained by simulating data of equivalent size on the basis of independent normal distribution. This quantile obtained for 40 individuals and 10 variables is about 38\%: although the percentage of 50\% seems relatively low, it expresses a significant structure in the data. However, the total variability cannot be summarized by the first two dimensions only. It may also be interesting to interpret components 3 and 4 for which the inertia is greater than 1 (this value is used as a reference because it represents, in the case of
standardized variables, the contribution of a single variable).
It can also be said that a high percentage means that the 10 events are redundant and that we do not need 10 sport events to rank athletes.


5) Comment: 
- the correlation between the 100 m and long.jump
- the correlation between long.jump and Pole.vault
- can you describe the athlete Casarsa? 
- the proximity between Sebrle and Clay 
- the proximity between Schoenbeck and Barras

- The 100m and long.jump are negatively correlated: therefore, an
athlete who runs 100 meters quickly will generally jump a long way. The
variables 100m, 400m, and 110m hurdles are positively correlated, that is, some
athletes perform well in all four events while others do not.

- Since long.jump is well represented in the first plan and Pole.vault is not, we
can deduce that long.jump and Pole.vault are approximately orthogonal, meaning
that the corresponding variables are roughly uncorrelated.

Overall, the variables relating to speed are negatively correlated with the
first principal component while the variables shot put and long jump are
positively correlated with this component. The coordinates of these active
variables can be found in the object res.pca$var which also gives the
representation quality of the variables (cosine squared) and their contributions
to the construction of the components.

```{r, echo = TRUE, eval = TRUE}
round(cbind(res.pca$var$coord[,1:4],res.pca$var$cos2[,1:4],
res.pca$var$contrib[,1:4]),2) 
```

Bourguignon and Karpov have very different performance profiles since they are
opposed according to the main axis of variability. 


- Casarsa is located on the top left corner. The first dimension is highly correlated with the number of points: this indicates that he does not have a large number of points. The second dimension is correlated with the Shot.put, High.jump and Discus. This indicates that Casarsa had good results in these three sports. Remember that the second dimension is calculated orthogonally to the first. So Casarsa has good results in these three sports compared to other "bad" athletes.

- Sebrle and Clay are close to one another and both far from the center of gravity of the cloud of points. The quality of their projection is therefore good, and we can be certain that they are indeed close in the original space. This means that they have similar profiles in their results across all sports events.

- Schoenbeck and Barras are close to one another but they are also close to the center of gravity of the cloud of points. When looking at their cos2 they are not well projected, We cannot interpret their distance based on this plot only.


The coordinates of these individuals can be found in
res.pca$ind. Below is a sample output provided by the PCA function. We obtain a
data table with the coordinates, the cosine squared (which gives an idea of the
quality of the projection of the individuals on a component), and the
contributions for each individual (to know how much an individual contributes to
the construction of a component).

```{r, echo = TRUE, eval = TRUE}
round(cbind(res.pca$ind$coord[,1:4],res.pca$ind$cos2[,1:4],
res.pca$ind$contrib[,1:4]),2) 
```

6) Enhance the graphical outputs with the following options:
```{r eval = TRUE}
 plot.PCA(res.pca, choix = "ind", habillage = ncol(decathlon), cex = 0.7)
 plot.PCA(res.pca, choix = "ind", habillage = ncol(decathlon), cex = 0.7, 
 autolab = "no")
 plot(res.pca, select = "cos2 0.8",  invisible = "quali")
 plot(res.pca, select = "contrib 10")
 plot(res.pca, choix = "var", select = "contrib 8", unselect = 0)
 plot(res.pca, choix = "var", select = c("400m", "1500m"))
```

7) In which trials those who win the decathlon perform the best? Could we say that the decathlon trials are well selected? 

The supplementary variable "number of points" is almost collinear to the first
principal component. Therefore, the athletes with a high number of points are
particularly good in the trials correlated with the first principal component.
Those who win the decathlon perform the best in 100m, 110m hurdles and long jump. This means that the ranking of the decathlon is governed by those three sports.


8) Compare and comment the performances during both events: Decastar and Olympic. Could we conclude on the differences?  Perform a test or plot Confidence ellipses:

```{r }
dimdesc(res.pca)
dimdesc(res.pca, proba = 0.2) 
plotellipses(res.pca)
```

This function is very useful when there are a great number of variables. We see
here that the first component is mainly due to the variable number of points
(with a correlation coefficient of 0.96), and the variable 100m (with a negative
correlation). The second component is described by two quantitative variables
only (discus and shot put). No category of any categorical variable
characterizes components 1 and 2 with a confidence level of 95%. 

However, with the confidence level of 0.8, we can say that both of the two categories
Olympic Games and Decastar have coordinates that are significantly different
from 0 on the first component. As the value is positive (negative) for the
Olympic Games (Decastar) we can say that individuals who participated in the
Olympic Games tended to have positive coordinates (or negative, respectively) on
component 1.

9) To select the number of dimensions, you should have a look at 
```{r eval = TRUE}
?estim_ncp
```
which performs cross validation as detailed in the lecture slides.

** More interpretation**

The representations of both the cloud of individuals and the cloud of variables 
are to be analysed together. In other words, differences between individuals can
be explained by the variables, and relationships between variables can be 
illustrated by individuals. On the whole, the first component opposes
performance profiles that are uniformly high" (i.e., athletes that are good in
all events) such as Karpov at the Olympics to performance profiles that are
(relatively!) weak in all events" such as Bourguignon at the Decastar meeting. 
Furthermore, the first component is mainly linked to the events using qualities
relating to a burst of energy (100m, 400m, 110m hurdles and long jump). These
four variables constitute a relatively homogeneous group: the correlation
between any two of these performances is higher than 0.52 (see the correlation
matrix). With just one exception, these variables have the highest coefficients.
This group of variables draws" (i.e., contributes to the construction of) the
first principal component and the overall score. It must here be emphasized that
the first principal component is the combination that best sums up all the
variables. In this example, the automatic summary provided by the PCA
corresponds almost exactly with the official summary (the number of points). The
second component opposes the variables of endurance (400m and 1500m) and power
(discus, shot put). Notably, it separates the performance profiles that are
considered weak", which suggests that the best performance profiles are
balanced: even among the weakest profiles, athletes can be spe- cialised. Note
that power (discus and shot put) is not correlated with speed (100m, long jump,
110m hurdles). As these two variables are not linearly related, there are
powerful and fast individuals (all-round athletes with high values on the first
component), powerful individuals who are not so fast (corresponding to high
values on com- ponent 2) and individuals who are not powerful but fast (with low
coordinates on component 2). This can be illustrated by comparing Casarsa and
Lorenzo from the stan- dardised data (see the table on the next page). Casarsa
performs well in the events that require power and poorly in speed related
events, while the oppo- site is true for Lorenzo. It must be noted that these
two athletes have a low coordinate on the first principal component and
therefore do not have good overall performances. Their strengths, which are
lessened by the second com- ponent, must therefore be relativized compared to
their overall performance. The variable number of points seems to be entirely
unrelated to this component (correlation of 0.02, see list of coordinates of
the supplementary variables). The third component is mainly related to the 1500
meters and to a lesser extent, to the pole vault. It opposes these two events:
athletes that do not perform well in the 1500 metres (N.B. as it is a variable
relating to speed, a high value indicates rather poor performance) do however
obtain good results in the pole vault (i.e., see standardized values in the
centered and scaled data for Terek: 1.96 in the pole vault and 0.98 in the 1500
meters). This third component mainly highlights four individuals that are
particularly weak in the 1500 meters: Clay and Karpov at Decastar (with
standardised val- ues of 1.95 and 1.84, respectively), and Terek and Kokhizoglou
at the Olympic Games (with standardised values of 0.98 and 3.29). These four
individuals contribute up to 34.7% of the inertia of component 3. The fourth
component is correlated with the variable javelin and, to a lesser extent, the
variable pole vault. Three profiles are characterized by these two events:
Bernard at the Decastar meeting, and Sebrle and Nool at the Olympic Games. These
three athletes contribute up to 31.3% of the inertia of this component. It must
be noted that the representations of the individuals and variables are only
approximate representations of the data table on the one hand, and of the
correlation (or variance{covariance) matrix on the other. It is therefore 
necessary to support the interpretation by referring back to the data  by looking at the means
the standard deviations by variable, the standardized data, and the correlation
matrix.

All athletes who participated in both decathlons certainly focused their phys-
ical preparation on their performances at the Olympic Games. Indeed, they
all performed better at the Olympic Games than at the Decastar meeting.
We can see that the dots representing a single athlete (for example, Sebrle)
are in roughly the same direction. This means, for example, that Sebrle is
good at the same events for both decathlons, but that the dot corresponding
to his performance at the Olympic Games is more extreme, so he obtained
more points during the Olympics than at the Decastar meeting.
This data can be interpreted in two different ways:
1. Athletes that participate in the Olympic Games perform better (on
average) than those participating in the Decastar meeting.
2. During the Olympics, athletes are more motivated by the challenge,
they tend to better, etc.
From the point of view of component 2, however, there is no overall
difference between the Olympics and Decastar. Overall, athletes' perfor-
mances may have improved but their profiles have not changed. Only Zsivoczky
changed from a rather powerful profile to a rather fast profile. His per-
formances in shot put and javelin are worse at Decastar compared to the
Olympics, with throws of 15.31m (standardised value of 1.02) and 13.48 (1:22) for shot put and 63.45m (1.08) and 55.37m (0:62) for javelin. However, he improved his performance in the 400 meters: 49.40 seconds and then 48.62 seconds and for the 110m hurdles: 14.95 seconds and then 14.17 seconds